---
title: "Daily Activity Analysis"
author: "Charlie Kriefall"
date: "March 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This document will demonstrate the exercises using a knitr document.
Code is embedded using proper markup symbols

1. Loading and preprocessing the data

```{r}
# 1. Loading and preprocessing the data
setwd("~/RData")

# create working directory
if (!dir.exists("activity")) dir.create("activity")
# go to working directory
setwd("activity")
# download dataset
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",destfile="activity.zip")
# unzip the file
unzip("activity.zip",overwrite=TRUE)

# load csv into a data table
library(readr)
activity <- read_csv("~/RData/activity/activity.csv", col_types = cols(date = col_date(format = "%Y-%m-%d")))


```

2. Create histogram of steps per day

```{r}
hist(aggregate(steps ~ date,data=activity,sum)$steps,main="Histogram Steps Per Day",xlab="steps")

```

3. mean and median of the total number of steps taken per day

```{r}
# mean steps per day
mean(aggregate(steps ~ date,data=activity,sum)$steps)
# median steps per day
median(aggregate(steps ~ date,data=activity,sum)$steps)

```

4. time series plot interval vs average steps

```{r}
plot(aggregate(steps ~ interval,data=activity,mean)$interval,aggregate(steps ~ interval,data=activity,mean)$steps,type="l",xlab="Interval",ylab="Steps",main="Mean Steps per Interval")


```

5. identify interval with max steps

```{r}
# create summary table of mean steps per interval
meanStepsPerInterval<-aggregate(steps ~ interval,data=activity,mean)
subset(meanStepsPerInterval, steps == max(meanStepsPerInterval$steps), select = c(interval))

```

6. Imputing missing values

Strategy: Replace NA steps with the mean value for the interval

```{r}
# add column meanSteps to dataset activity by joining to derived dataset meanStepsPerInterval
activity$meanSteps <- meanStepsPerInterval[match(activity$interval,meanStepsPerInterval$interval),2]
# update the NA values in steps with the value in meanSteps
activity$steps[is.na(activity$steps)] <- activity$meanSteps[is.na(activity$steps)]

```

7. Histogram of total steps per day with imputed values

```{r}
# add a column called weekday based on date
activity$weekday <- weekdays(activity$date)
par(mfrow=c(1,1))
hist(aggregate(steps ~ weekday,data=activity,sum)$steps,main="Histogram of Steps Per Day")

```

8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}
activity$IsWeekend <- activity$weekday=="Saturday"|activity$weekday=="Sunday"

# side by side panels
par(mfrow=c(1,2))
# weekend
plot(aggregate(steps ~ interval,data=activity[activity$IsWeekend,],mean)$interval,aggregate(steps ~ interval,data=activity[activity$IsWeekend,],mean)$steps,type="l",xlab="5 minute interval",ylab="steps",main="Weekend Steps")
# weekdays
plot(aggregate(steps ~ interval,data=activity[!activity$IsWeekend,],mean)$interval,aggregate(steps ~ interval,data=activity[!activity$IsWeekend,],mean)$steps,type="l",xlab="5 minute interval",ylab="",main="Weekday Steps")


```

Clean Up R Environment

```{r}
setwd("~/RData")
rm(list=ls())
par(mfrow=c(1,1))
```

